---
title: "Human Activity Recognition - A Prediction with Wearables"
author: "thenuv"
date: "3 August 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This is a study with data on Human activity recognition. A prediction is made on the way in which the exercise was performed. There are 2 datasets for Training and Test. Tree based modelling was built for the training set and a subset of it is used to evaluate the prediction. After comparing the models, the chosen one is used to predict the test set.


## Data Processing

```{r DataProcess}
# Load Libraries
suppressMessages(suppressWarnings(library(caret)))
suppressMessages(suppressWarnings(library(randomForest)))
suppressMessages(suppressWarnings(library(rpart)))
suppressMessages(suppressWarnings(library(rpart.plot)))

# Download Data
fileurl1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
fileurl2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
filename1 <- "pml_training.csv"
filename2 <- "pml_testing.csv"
if (!file.exists(filename1)) {
        download.file(fileurl1, filename1)
}

if (!file.exists(filename2)) {
        download.file(fileurl2, filename2)
}
# Import Data
df_train <- read.csv(filename1) 
df_test <- read.csv(filename2) 

```

## Cleanse & Explore 

As part of cleansing and filtering, the following approach is used to remove some variables as they may not contribute as predictors in the model to be trained.

-   Filter near Zero Variance Columns
-   Misc. columns like Name, X & timestamp
-   Columns having NA's.

``` {r Explore}
dim(df_train)
dim(df_test)
#str(df_train)
#str(df_test)
table(df_train$classe)

# Filter near Zero Variance columns
NZV <- nearZeroVar(df_train, saveMetrics = TRUE)
#head(NZV,10)
df_train <- df_train[,!NZV$nzv] 
df_test <- df_test[,!NZV$nzv]

# Drop columns which won't be used in the study
drops <- grepl("^X|user_name|timestamp", names(df_train))
df_train <- df_train[,!drops] 
df_test <- df_test[,!drops]

# Retain columns not having NA's
nonnacol <- (colSums(is.na(df_train)) == 0)
df_train <- df_train[, nonnacol] 
df_test <- df_test[, nonnacol] 

dim(df_train)
dim(df_test)

```



```{r ModelPrep}

# Split training data into training and validation
set.seed(2345)
inTrain <- createDataPartition(df_train$classe, p=.70, list=FALSE)
training <- df_train[inTrain, ]
validation <- df_train[-inTrain, ]
dim(training)
dim(validation)

```

### Model 1 with rpart approach
 Tree based rpart model is built on training set as bellow.
```{r MD1 }
md1 <- rpart(classe ~. , data=training, method="class")
prp(md1)

```

Predict Model 1 with Validation data set.
```{r ValidateMD1}

predict1 <- predict(md1, validation, type="class")
confusionMatrix (validation$classe, predict1)

accuracy_md1 <- postResample(predict1, validation$classe)[1]
se_md1 <- 1- as.numeric(confusionMatrix(validation$classe, predict1)$overall[1])

```
The estimated *accuracy* for this model is ***74.71%*** with expected out of sample error ***25.29%***.

### Model 2 Random Forest approach
 We apply the Random Forest method to train the data. Cross validation with 5 fold opton is specified in the traning control parameters.
 
``` {r MD2}
md2 <- train(classe~., data=training, method="rf", trControl = trainControl(method="cv", 5), ntree=50)
md2


```

Predict with  Model2 on validation data set.
```{r Validate}
# Validate
predict2 <- predict(md2, validation)
confusionMatrix(validation$classe, predict2)
accuracy_md2 <- postResample(predict2, validation$classe)[1]
se_md2 <- 1- as.numeric(confusionMatrix(validation$classe, predict2)$overall[1])

```
The outcome accuracy is estimated at ***99.69%*** with sample error of ***0.31% ***. 

## Predict Test Data
Since the estimated outcome accuracy of Random Forest approach is better than the other model, we use it to predict the test data set. 
``` {r PredictTest}
# Predict Test
predictTest <- predict(md2, df_test)
predictTest
```


### Reference
http://groupware.les.inf.puc-rio.br/har
